\chapter*{Abstract}

Web server optimization is important for businesses since web servers can achieve higher performance by consuming fewer resources. This research presents optimization techniques for different program types based IO characteristics. Several web server architectures are implemented in Ballerina and then tuned. Performance of each server architecture and tunes are evaluated against different types of programs implemented where CPU and IO intensive features are prominent in each program. For further, tuning thread pool size is selected based on experimental results. Then optimal thread pool size is identified for each set of programming features where features are different types of remote IO calls. Since IO calls are part of the native representation of ballerina, a parser is implemented to extract these features automatically from the Abstract Syntax Tree of the Ballerina source code.  Experiments are run on Google Cloud VM and docker containers were used to properly isolate and resource controlling for benchmark. Jmeter is used as the primary tool for obtaining evaluation metrics.

A machine learning model is proposed where optimal thread pool size is the output of a given set of programming features. The decision tree Regression model performed best among other selected models. Finally, it could be observed that for different program types, the latency when the optimal thread pool size suggested by the ML approach is used is lower than the latency when Ballerina's default thread pool size is used.