\chapter*{Preface}

The purpose of this dissertation is to present the work from the research named "Machine learning approach to fine-tune Ballerina thread pool size". Final solution includes a machine learning model that estimate the thread pool size of Ballerina scheduler based on program features. Starting point of this study was comparing performance between different server architectures in Ballerina and evaluating performance for different types of Ballerina programs. Then discussed how to improve performance further for different programs especially for programs that consists of different types of IO features. It is identified thread pool optimization is the best way to improve the performance. Experiments were run in each phase and results were justified. Finally, it is shown that predicted thread pool size from the machine learning model is better than the current Ballerina architecture. There were no performance comparison done for different sever architectures and for different thread pool size for Ballerina language. Furthermore, no study were conduct to extract features from the source code and estimate the parameters that would give the best performance for a web server. 

My original works include implementing different server architectures for ballerina, performance comparison of each other, identification of different prgram features that affect the performance of web server, implementation of many different programs in order to train the machine learning model, and implementing a machine learning model that predict the optimal thread pool size.